---
title: "Training a new DLBCLone model"
vignette: >
  %\VignetteIndexEntry{quarto vignettes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
from: markdown+emoji
fig.align: "center"
---

```{r set working directory and libs, echo=FALSE, include=FALSE, cache=FALSE}
library(GAMBLR.predict)
library(tidyverse)

default_model <- "best_opt_w"

all_features_optimized <- DLBCLone_load_optimized(
        dirname(
            system.file(
                "extdata/models",
                "best_opt_w_model.rds",
                package = "GAMBLR.predict"
            )
        ),
        default_model
    )

```

## Preamble

If you do not have a DLBCLone model suitable for your application, you can train a custom model to meet your needs. The training process leverages bundled mutation data and LymphGen classifications for over 2000 DLBCLs. Once you have a model you're satisfied with, it can be applied to your own data using `DLBCLone_predict`.  

## Optimizing parameters

The `DLBCLone_optimize_params` function tunes the K value and a few thresholds based on the training data, working to maximize balanced accuracy or (optionally) another metric. The required and commonly used arguments are:

* `combined_mutation_status_df` feature matrix data frame with one row per sample and one column per mutation


Confusingly, this argument is no longer required. The mutation status is stored in the output of `make_and_annotate_umap`

* `metadata_df` data frame of metadata with one row per sample and three required columns: sample_id, dataset and the column containing the "truth" lable, which defaults to lymphgen


* `umap_out` the output of `make_and_annotate_umap`

* `truth_classes` vector of classes to use for training and testing (LymphGen default: c("EZB", "MCD", "ST2", "N1", "BN2", "Other"))

* `min_k` lowest k in the range of values to try for KNN (default: 3). Set higher to restrict to a narrower range. 

* `max_k` highest k in the range of values to try for KNN (default: 33). Set lower to restrict to a narrower range. 


## Example

If you're here, then we assume you've run `make_and_annotate_umap` and are generally happy with the separation of samples in this latent representation. We can use `DLBCLone_optimize_params` to automate finding an optimal K value (number of neighbors) and a few thresholds to categorize each sample based on its location and proximity to other points in the UMAP. Here, you provide the original UMAP produced by `make_and_annotate_umap` and *not* the re-projected UMAP, because this function will perform the sample-by-sample projection for us. 

```{r optimize_params, eval=FALSE}

best_opt_model = DLBCLone_optimize_params(  
  combined_mutation_status_df = mu_everything$features, 
  metadata_df = dlbcl_meta_clean,
  umap_out = mu_everything,
  truth_classes = c("MCD","EZB","BN2","N1","ST2","Other"),
  min_k=5,
  max_k=13
)
```


## Overview of classifications

### make_alluvial 

There is a convenience plotting function `make_alluvial` generates alluvial plots that summarize the relationship between original (e.g., Lymphgen) and predicted (e.g., DLBCLone) class assignments for a set of samples. Typically, you would use these to evaluate how well your model performs across each of the classes. The only required argument is the DLBCLone model, for example, the output of `DLBCLone_optimize_params`. Some additional arguments that are commonly used:

* `pred_column` name of the column in predictions to use for the predicted class. This is usually one of DLBCLone_wo or DLBCLone_w, respectively for DLBCLone with or without outgroup optimization 

* `pred_name` Display name for the predictions (e.g. DLBCLone)


```{r plot_alluvial1, message=FALSE, warning=FALSE,results='hide'}
make_alluvial(all_features_optimized,
  pred_column = "DLBCLone_wo",
  pred_name = "DLBCLone Optimized")
```

```{r plot_alluvial2, message=FALSE, warning=FALSE,results='hide'}
make_alluvial(all_features_optimized,
  pred_column = "DLBCLone_w",
  pred_name = "DLBCLone Greedy")
```

### make_umap_scatterplot

An overview of the classification result can also be visualized in our UMAP space. It can be useful to compare the UMAP coloured the ground truth labels to the new labels produced by DLBCLone.

```{r scatter1, message=FALSE, warning=FALSE,results='hide'}
make_umap_scatterplot(
  all_features_optimized$prediction, 
  colour_by = "lymphgen",
  title = "Ground Truth"
)
```

```{r scatter2, message=FALSE, warning=FALSE,results='hide'}
make_umap_scatterplot(
  all_features_optimized$prediction, 
  colour_by = "DLBCLone_wo",
  title = "DLBCLone Optimized"
)
```


