---
title: "DLBCLone Functions - DLBCLone Model Management"
vignette: >
  %\VignetteIndexEntry{quarto vignettes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
from: markdown+emoji
fig.align: "center"
---

# Save and Load Optimize Params 

For future use cases and reproducibility purposes you may store the outputs of `DLBCLone_optimize_params`.

## DLBCLone_save_optimized

Allows you to store your model in a given directory with a name prefix.

```{r, eval=FALSE}
DLBCLone_save_optimized( # saving DLBCLone_optimize_params
  DLBCLone_model = best_opt_model,
  base_path="../data/saved_tutorial_model",
  name_prefix="best_opt_model"
)
```

## DLBCLone_load_optimized

Allows you to load in your stored model.

```{r, eval=FALSE}
loaded_model <- DLBCLone_load_optimized( # loading DLBCLone_optimize_params
   path="../data/saved_tutorial_model",
   name_prefix="best_opt_model"
)
```

## DLBCLone_activate

Reactivates a model after optimization, setting the UMAP projections.

```{r, eval=FALSE}
best_opt_model <- DLBCLone_activate(loaded_model, force = TRUE)
```