---
title: "Gaussian Mixture Model"
vignette: >
  %\VignetteIndexEntry{quarto vignettes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
from: markdown+emoji
fig.align: "center"
---

```{r, echo = FALSE, message=FALSE, warning=FALSE, results='hide'}
library(GAMBLR.predict)
library(tidyverse)
```

A variant of DLBCLone that doesn't use K-nearest neighbors relies instead on a gaussian mixture model and UMAP embeddings. This functionality is experimental. It’s important not to confuse this modeling with the previously introduced DLBCLone functions.

# DLBCLone_train_mixture_model 

Training a Gaussian Mixture Model for DLBCLone Classification, this function fits a supervised Gaussian mixture model (GMM) to UMAP-projected data for DLBCLone subtypes, excluding samples labeled "Other". It assigns class predictions and optionally reclassifies samples as "Other" based on probability and density thresholds.

`umap_out` output from `make_and_annotate_umap`, containing a data frame with UMAP coordinates and truth labels

`probability_threshold` minimum posterior probability required to assign a class (default: 0.5)

`density_max_threshold` minimum maximum density required to assign a class (default: 0.05)

```{r, eval=FALSE}
library(mclust) # <- necessary for running the mixture models

mixture_result <- DLBCLone_train_mixture_model(umap_out = mu_everything)
```

# DLBCLone_predict_mixture_model 

Applies a previously trained supervised Gaussian mixture model (GMM) to UMAP-projected data for DLBCLone subtypes. Assigns class predictions and optionally reclassifies samples as "Other" based on probability and density thresholds.

`model` fitted model object, as returned by `DLBCLone_train_mixture_model`

`umap_out` output from `make_and_annotate_umap`, containing a data frame with UMAP coordinates for the samples to be classified with the model. This must be projected using the same UMAP model that was generated using the training data.

`probability_threshold` minimum posterior probability required to assign a class (default: 0.5)

`density_max_threshold` minimum maximum density required to assign a class (default: 0.05)

```{r, eval=FALSE}
predict_mixture_result <- DLBCLone_predict_mixture_model(
  model = mixture_result$gaussian_mixture_model, 
  umap_out = mu_everything
)
```
