---
title: "DLBCLone Full Walkthrough of Associated Functions"
vignette: >
  %\VignetteIndexEntry{quarto vignettes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
from: markdown+emoji
fig.align: "center"
---

```{r set working directory and libs, echo=FALSE, include=FALSE, cahce=FALSE}
library(GAMBLR.predict)
library(GAMBLR.helpers) 
library(tidyverse)

default_model <- "best_opt_w"

loaded_model <- DLBCLone_load_optimized(
        dirname(
            system.file(
                "extdata/models",
                "best_opt_model.rds",
                package = "GAMBLR.predict"
            )
        ),
        default_model
    )

best_opt_model <- DLBCLone_activate(loaded_model, force = TRUE)
```

*A step by step example workflow from start to finish*

```{r, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(GAMBLR.open)
library(tidyverse)
```

## Load the GAMBLR.predict package associated training files from extdata: 
```{r, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
best_opt_model_genes <- sort(c(
  "ACTB", "ACTG1", "BCL10", "BCL2", "BCL2L1", 
  "BCL6", "BIRC3", "BRAF", "BTG1", "BTG2", 
  "BTK", "CD19", "CD70", "CD79B", "CD83", 
  "CDKN2A", "CREBBP", "DDX3X", "DTX1", "DUSP2", 
  "EDRF1", "EIF4A2", "EP300", "ETS1", "ETV6", 
  "EZH2", "FAS", "FCGR2B", "FOXC1", "FOXO1", 
  "GNA13", "GRHPR", "HLA-A", "HLA-B", "HNRNPD", 
  "IRF4", "IRF8", "ITPKB", "JUNB", "KLF2", 
  "KLHL14", "KLHL6", "KMT2D", "MEF2B", "MPEG1", 
  "MYD88", "NFKBIA", "NFKBIE", "NFKBIZ", "NOL9", 
  "NOTCH1", "NOTCH2", "OSBPL10", "PIM1", "PIM2", 
  "PRDM1", "PRDM15", "PRKDC", "PRRC2C", "PTPN1", 
  "RFTN1", "S1PR2", "SETD1B", "SGK1", "SOCS1", 
  "SPEN", "STAT3", "STAT6", "TBCC", "TBL1XR1", 
  "TET2", "TMEM30A", "TMSB4X", "TNFAIP3", "TNFRSF14", 
  "TOX", "TP53", "TP73", "UBE2A", "WEE1", 
  "XBP1", "ZFP36L1", "MYD88HOTSPOT", "BCL2_SV", "BCL6_SV"
))

best_opt_model_matrix <- readr::read_tsv(system.file("extdata/all_full_status.tsv",package = "GAMBLR.predict")) %>%
  tibble::column_to_rownames("sample_id") %>%
  select(best_opt_model_genes)

dlbcl_meta <- readr::read_tsv(system.file("extdata/dlbcl_meta_with_dlbclass.tsv",package = "GAMBLR.predict"))

dlbcl_meta_clean <- filter(
  dlbcl_meta,
  lymphgen %in% c("MCD","EZB","BN2","N1","ST2","Other")
)
```

## stacked_bar_plot

```{r, message=FALSE, warning=FALSE}
plots <- posthoc_feature_enrichment(
  sample_metadata = dlbcl_meta_clean,
  features = best_opt_model_matrix,
  method = "fisher",
  num_feats = 15,
  title = "LymphGen"
)

plots$bar_plot
```

## forest_plot

```{r}
#| fig-height: 12
#| fig-width: 4
plots$forest_plot
```

## make_and_annotate_umap
```{r umap1, message=FALSE, warning=FALSE,results='hide'}
mu_everything = make_and_annotate_umap(
  best_opt_model_matrix,
  dlbcl_meta_clean,
  core_features = list(
    ST2=c("SGK1","DUSP2","TET2","SOCS1"),
    N1="NOTCH1",
    EZB=c("EZH2","BCL2_SV"),
    MCD=c("MYD88HOTSPOT","CD79B","PIM1"),
    BN2=c("BCL6_SV","NOTCH2","SPEN","CD70")
  )
)
```

## make_umap_scatterplot 

```{r}
make_umap_scatterplot(
  mu_everything$df,
  drop_other = F
)
```

## DLBCLone_optimize_params 

```{r optimize1, eval = FALSE}
best_opt_model = DLBCLone_optimize_params(  
  mu_everything$features, 
  dlbcl_meta_clean,
  umap_out = mu_everything,
  truth_classes = c("MCD","EZB","BN2","N1","ST2","Other"),
  optimize_for_other = T,
  min_k=5,
  max_k=13
)
```

## make_alluvial

DLBCLone with outgroup optimization ("wo") w for weighted, o for optimized for other.
```{r plot_alluvial2, message=FALSE, warning=FALSE,results='hide'}
make_alluvial(best_opt_model,pred_column = "DLBCLone_wo",pred_name = "DLBCLone_wo")
```

## DLBCLone_predict

### predicting one training sample
```{r, message=FALSE, warning=FALSE}
test_sample = "00-14595_tumorC"

pred_train <- DLBCLone_predict(
  mutation_status = best_opt_model_matrix[test_sample,],
  optimized_model = best_opt_model
)
knitr::kable(head(pred_train$prediction))
```

## make_neighborhood_plot 
```{r, message=FALSE, warning=FALSE}
make_neighborhood_plot(
  single_sample_prediction_output = pred_train,
  this_sample_id = "00-14595_tumorC",
  prediction_in_title = TRUE,
  add_circle = TRUE,
  label_column = "DLBCLone_wo"
)
```

## nearest_neighbor_heatmap 

```{r, message=FALSE,warning=FALSE}
nearest_neighbor_heatmap(
  this_sample_id = "00-14595_tumorC",
  DLBCLone_model = pred_train,
  font_size = 7
)
```
