---
title: "Predicting classes with DLBCLone_predict"
vignette: >
  %\VignetteIndexEntry{quarto vignettes}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
from: markdown+emoji
fig.align: "center"
---

# Predict Samples

```{r set working directory and libs, echo=FALSE, include=FALSE, cahce=FALSE}
library(GAMBLR.predict)
library(GAMBLR.helpers) 
library(tidyverse)

best_opt_model_genes <- sort(c(
  "ACTB", "ACTG1", "BCL10", "BCL2", "BCL2L1", 
  "BCL6", "BIRC3", "BRAF", "BTG1", "BTG2", 
  "BTK", "CD19", "CD70", "CD79B", "CD83", 
  "CDKN2A", "CREBBP", "DDX3X", "DTX1", "DUSP2", 
  "EDRF1", "EIF4A2", "EP300", "ETS1", "ETV6", 
  "EZH2", "FAS", "FCGR2B", "FOXC1", "FOXO1", 
  "GNA13", "GRHPR", "HLA-A", "HLA-B", "HNRNPD", 
  "IRF4", "IRF8", "ITPKB", "JUNB", "KLF2", 
  "KLHL14", "KLHL6", "KMT2D", "MEF2B", "MPEG1", 
  "MYD88", "NFKBIA", "NFKBIE", "NFKBIZ", "NOL9", 
  "NOTCH1", "NOTCH2", "OSBPL10", "PIM1", "PIM2", 
  "PRDM1", "PRDM15", "PRKDC", "PRRC2C", "PTPN1", 
  "RFTN1", "S1PR2", "SETD1B", "SGK1", "SOCS1", 
  "SPEN", "STAT3", "STAT6", "TBCC", "TBL1XR1", 
  "TET2", "TMEM30A", "TMSB4X", "TNFAIP3", "TNFRSF14", 
  "TOX", "TP53", "TP73", "UBE2A", "WEE1", 
  "XBP1", "ZFP36L1", "MYD88HOTSPOT", "BCL2_SV", "BCL6_SV"
))

best_opt_model_matrix <- readr::read_tsv(system.file("extdata/all_full_status.tsv",package = "GAMBLR.predict")) %>%
  tibble::column_to_rownames("sample_id") %>%
  select(best_opt_model_genes)

default_model <- "best_opt_w"

loaded_model <- DLBCLone_load_optimized(
        dirname(
            system.file(
                "extdata/models",
                "best_opt_model.rds",
                package = "GAMBLR.predict"
            )
        ),
        default_model
    )

all_features_optimized <- DLBCLone_activate(loaded_model, force = TRUE)
```

# DLBCLone_predict 

Now for the primary objective of DLBCLone. What happens when you have new samples with no metadata? Using the trained models produced, `DLBCLone_predict` can approximate labelings to samples based on their corresponding feature matrix. This function is versatile, you can use it to cross reference your training data, predict a test sample one at a time, or all of your test samples all at once! Here we will just be using training samples.

`mutation_status` data frame containing the mutation feature matrix you are testing on 

`optimized_model` list of parameters from DLBCLone_optimize_params

## predicting one training sample 
```{r predict single sample, message=FALSE, warning=FALSE}
test_sample = "00-14595_tumorC"

pred_train <- DLBCLone_predict(
  mutation_status = best_opt_model_matrix,
  optimized_model = all_features_optimized
)

knitr::kable(head(pred_train$prediction))
```

# make_neighborhood_plot 

Used for visualizing how and where neighbors are selected, a umap plot is produced showcasing the test sample of interest and its trained neighbors that helped determine its labeling.

`single_sample_prediction_output` output list from predict_single_sample_DLBCLone

`training_predictions` training predictions from DLBCLone_optimize_params (e.g. optimized_model$df)

`this_sample_id Character` the sample ID for which the neighborhood plot will be generated

`prediction_in_title` if TRUE: includes the predicted label in the plot title

`add_circle` Plot will include a circle surrounding the set of neighbors, helpful for identification

```{r}
make_neighborhood_plot(
  single_sample_prediction_output = pred_train,
  this_sample_id = "00-14595_tumorC",
  prediction_in_title = TRUE,
  add_circle = TRUE,
  label_column = "DLBCLone_wo"
)
```

# nearest_neighbor_heatmap 

Generates a heatmap of feature values for the nearest neighbors of a specified sample, based on a DLBCLone model object. This visualization helps to inspect the feature profiles of samples most similar to the query sample.

`this_sample_id` the sample ID for which to plot the nearest neighbor heatmap

`DLBCLone_model` a DLBCLone model object, which can be the output of DLBCLone_optimize_params, DLBCLone_KNN, or DLBCLone_predict

`truth_column` the column name in the predictions data frame that contains the ground truth labels (default: "lymphgen")

`metadata_cols` optional character vector of additional metadata columns to include in the heatmap annotations

`clustering_dis` the distance metric to use for clustering rows (default: "binary")

`font_size` font size for heatmap text (default: 14)

```{r, message=FALSE,warning=FALSE}
nearest_neighbor_heatmap(
  this_sample_id = "00-14595_tumorC",
  DLBCLone_model = pred_train,
  font_size = 7
)
```
